Static Race Detection in OS Kernels by Mining Locking Rules

### 2023-07-25 2.motivation.tex
update technique

\begin{figure}[htbp]
	\centering
	\includegraphics[width=1\linewidth]{figures/fig_bug_demo.pdf}
	\figcaption{A null-pointer dereference due to data race in Linux 6.2.}
	\label{fig_bug_demo}
\end{figure}
Figure~\ref{fig_bug_demo} shows a real null-pointer dereference caused by a 
data race in the Linux DRM driver. In the DRM driver, the functions {\tt 
exynos\_drm\_crtc\_atomic\_disable()} and {\tt exynos\_crtc\_handle\_event()} 
can execute concurrently. We exploit \textcircled{\footnotesize{n}} to 
represent the execution order of instructions and show one execution case in 
the left of Figure~\ref{fig_bug_demo}. In Thread 1, the variables {\tt 
crtc->state->event} and {\tt crtc->state->active} is checked by an if statement 
in the function {\tt exynos\_drm\_crtc\_atomic\_disable()}. After the condition 
is calculated to be true, the function {\tt exynos\_crtc\_handle\_event()} is 
executed in Thread 2. In this function, the value of {\tt crtc->state->active} 
is assigned to {\tt event} (\textcircled{\footnotesize{2}}) and then {\tt 
event} is checked in an if statement (\textcircled{\footnotesize{3}}). If it is 
not NULL, the variable {\tt crtc->state->event} is assigned with NULL 
(\textcircled{\footnotesize{4}}). Right after this assignment, the function 
{\tt drm\_crtc\_send\_vblank\_event} is called  in Thread 1 
(\textcircled{\footnotesize{6}}) with the argument {\tt crtc->state->active}, 
after acquiring the lock {\tt crtc->dev->event\_lock} 
(\textcircled{\footnotesize{5}}). In the called function, the variable {\tt 
crtc->state->event} is dereferenced through {\tt e->pipe} 
(\textcircled{\footnotesize{6}}). In this execution case, the data structure 
{\tt crtc->state->event} is first assigned with NULL and then dereferenced, and 
thus a null-pointer dereference can occur.

>>>>

\begin{figure}[htbp]
	\centering
	\includegraphics[width=1\linewidth]{figures/fig_demo_bug.pdf}
	\figcaption{A null-pointer dereference due to data race in Linux 6.2.}
	\label{fig_demo_bug}
\end{figure}
Figure~\ref{fig_demo_bug} shows a real null-pointer dereference caused by a 
data race in the Linux DRM driver. In the DRM driver, the functions {\tt 
exynos\_drm\_crtc\_atomic\_disable()} and {\tt exynos\_crtc\_handle\_event()} 
can execute concurrently. We exploit \textcircled{\footnotesize{n}} to 
represent the execution order of instructions and show one execution case in 
the left of Figure~\ref{fig_demo_bug}. In Thread 1, the variables {\tt 
crtc->state->event} and {\tt crtc->state->active} is checked by an if statement 
in the function {\tt exynos\_drm\_crtc\_atomic\_disable()}. After the condition 
is calculated to be true, the function {\tt exynos\_crtc\_handle\_event()} is 
executed in Thread 2. In this function, the value of {\tt crtc->state->active} 
is assigned to {\tt event} (\textcircled{\footnotesize{2}}) and then {\tt 
event} is checked in an if statement (\textcircled{\footnotesize{3}}). If it is 
not NULL, the variable {\tt crtc->state->event} is assigned with NULL 
(\textcircled{\footnotesize{4}}). Right after this assignment, the function 
{\tt drm\_crtc\_send\_vblank\_event} is called  in Thread 1 
(\textcircled{\footnotesize{6}}) with the argument {\tt crtc->state->active}, 
after acquiring the lock {\tt crtc->dev->event\_lock} 
(\textcircled{\footnotesize{5}}). In the called function, the variable {\tt 
crtc->state->event} is dereferenced through {\tt e->pipe} 
(\textcircled{\footnotesize{6}}). In this execution case, the data structure 
{\tt crtc->state->event} is first assigned with NULL and then dereferenced, and 
thus a null-pointer dereference can occur.

### 2023-07-25 3.technique.tex
update technique

\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/fig_alias_graph_demo.pdf}
	\figcaption{Example of alias graph.}
	\label{fig_alias_graph_demo}
\end{figure}

\noindent{\textbf{\em Example.}} Figure~\ref{fig_alias_graph_demo} shows a 
piece of driver-like source code and its alias graph. In this example, after an 
GEP (\&dev->lock) operation at Line 6, an edge labeled with {\tt lock} from 
node $\mathit{n_1}$ to node $\mathit{n_4}$ is inserted. Similarly, an edge 
labeled with {\tt flag} from node $\mathit{n_1}$ to node $\mathit{n_2}$ is 
inserted after Line 7. At last, an edge labeled with a dereference operator 
($\mathit{*}$) is inserted after the STORE (*fp = f) operation at Line 8. The 
final alias graph is shown in Figure~\ref{fig_alias_graph_demo}(b), and access 
paths are shown in the bottom left corner. Take node $\mathit{n_3}$ as an 
example, it can represent two fields, one is {\tt Dev.flag*}, and the other is 
{\tt int\_pointer*} (we exploit int\_pointer to represent a pointer points to 
an integer, and regard it as a data structure for convenience).

@@

Take the alias graph in Figure~\ref{fig_alias_graph_demo} as an example, {\tt 
\&dev->flag} is represented by node $\mathit{n_2}$, and {\tt \&dev->lock} is 
represented by node $\mathit{n_4}$. The two nodes have a common ancestor 
$\mathit{n_1}$, and thus the accessed variable {\tt \&dev->flag} and the 
protecting lock {\tt \&dev->lock} can be inferred to exist in the same data 
structure (namely {\tt Dev}). Therefore, the structure field Dev.flag is likely 
to be protected the lock stored in the structure field Dev.lock.

@@

>>>>

\begin{figure}[htbp]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/fig_demo_alias_graph.pdf}
	\figcaption{Example of alias graph.}
	\label{fig_demo_alias_graph}
\end{figure}

\noindent{\textbf{\em Example.}} Figure~\ref{fig_demo_alias_graph} shows a 
piece of driver-like source code and its alias graph. In this example, after an 
GEP (\&dev->lock) operation at Line 6, an edge labeled with {\tt lock} from 
node $\mathit{n_1}$ to node $\mathit{n_4}$ is inserted. Similarly, an edge 
labeled with {\tt flag} from node $\mathit{n_1}$ to node $\mathit{n_2}$ is 
inserted after Line 7. At last, an edge labeled with a dereference operator 
($\mathit{*}$) is inserted after the STORE (*fp = f) operation at Line 8. The 
final alias graph is shown in Figure~\ref{fig_demo_alias_graph}(b), and access 
paths are shown in the bottom left corner. Take node $\mathit{n_3}$ as an 
example, it can represent two fields, one is {\tt Dev.flag*}, and the other is 
{\tt int\_pointer*} (we exploit int\_pointer to represent a pointer points to 
an integer, and regard it as a data structure for convenience).

@@

Take the alias graph in Figure~\ref{fig_demo_alias_graph} as an example, {\tt 
\&dev->flag} is represented by node $\mathit{n_2}$, and {\tt \&dev->lock} is 
represented by node $\mathit{n_4}$. The two nodes have a common ancestor 
$\mathit{n_1}$, and thus the accessed variable {\tt \&dev->flag} and the 
protecting lock {\tt \&dev->lock} can be inferred to exist in the same data 
structure (namely {\tt Dev}). Therefore, the structure field Dev.flag is likely 
to be protected the lock stored in the structure field Dev.lock.

@@

\begin{figure}[htbp]
	\centering
	\includegraphics[width=1\linewidth]{figures/fig_demo_rule_mining.pdf}
	\figcaption{Example of locking-rule mining.}
	\label{fig_demo_rule_mining}
\end{figure}

\noindent{\textbf{\em Example.}} We use an example in 
Figure~\ref{fig_demo_rule_mining} to illustrate how to mine locking rules 
with the three stages. The analysis first perform a path-based analysis to 
extract key fields. Take the code path {\tt Line 12, Line 6, Line 7, Line 8, 
Line 9} as an example. The analysis updates the alias graph by handling two GEP 
operations ({\em \&dev->lock} and {\em fp = dev->flag}) and a STORE operation 
({\em *fp = f}). The final alias graph is shown in 
Figure~\ref{fig_demo_rule_mining}(b). At the same time of alias analysis, the 
analysis records the acquired lock {\em Dev.lock} into the lock set after the 
lock instruction at Line 6. When analyzing the read instruction at Line 7, the 
analysis finds the node of the accessed field ($\mathit{n_2}$) and the node of 
the lock stored in the lock set ($\mathit{n_4}$) has the same ancestor 
($\mathit{n_1}$), and thus {\em Dev.flag} is a key field. Similarly, {\em 
Dev.flag*} is also a key field, due to the write instruction at Line 8.

After extracting key fields, the method collects all accesses to key fields 
with lock set analysis and the alias graph. Take the code path {\tt Line 15, 
Line 6, Line 7, Line 8, Line 9, Line 16 and Line 17} as an example, through a 
lock instruction at Line 6, the analysis records the acquired lock {\em 
Dev.lock} into the lock set. When analyzing the read instruction at Line 7, the 
analysis finds the node of the accessed field ($\mathit{n_2}$) and the node of 
the lock stored in the lock set ($\mathit{n_4}$) has the same ancestor 
($\mathit{n_1}$), and thus records a field access <{\em Dev.flag}, {\em 
Dev.lock}>. Similar to the read instruction at Line 7, the method also records 
a field access <{\em Dev.flag*}, {\em Dev.lock}> for the write instruction at 
Line 8. Then through an unlock operation at Line 9, the analysis removes the 
lock {\em Dev.lock} is removed from the lock set. As a result, the key fields 
{\em Dev.flag} and {\em Dev.flag*} are accessed without acquiring any lock, and 
thus the analysis records two field accesses <{\em Dev.flag}, {\em NULL}> and 
<{\em Dev.flag*}, {\em NULL}>. The final field accesses are shown in 
Figure~\ref{fig_demo_rule_mining}.

After collecting all field accesses, the analysis finds that the fields {\em 
Dev.flag} and {\em Dev.flag*} are accessed under three calling contexts, and 
two of them are protected by the lock {\em Dev.lock}. And thus the two fields 
{\em Dev.flag} and {\em Dev.flag*} is deduce to be protected by the lock {\em 
Dev.lock} (assume the threshold of the ratio locked\_access\_num / 
all\_access\_num is 0.6).

\subsection{Lock-Usage Analysis}
\label{subsec_lock_usage_analysis}
